Quantitative Estimates of Minimum Core Masses for Giant Planet Formation

### 2013-06-17 tex/draft_paper2/draft_paper2.tex
current paper versions

\title{Equation of State Effects on the Minimum Core Mass for Giant Planet Formation}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}
\author{Ruth A. Murray-Clay}
\affil{Harvard-Smithsonian Center for Astrophysics}

@@

The core accretion model assumes that giant planets form through gas accretion on to a solid core. The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium; once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences. The minimum core mass for which unstable atmosphere collapse occurs is typically called the ``critical core mass''. In standard calculations of the critical core mass, planetesimal accretion dominates the atmosphere evolution, and the energy deposited by incoming planetesimal on to the core is radiated away by the atmosphere. In this study we consider a low planetesimal accretion regime in which the luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction. We use the atmosphere structure and cooling model developed in Piso \& Youdin (in prep.) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state. We find that the minimum core mass (which we denote as \textit{critical core mass}) to form a giant planet before the dissipation of the protoplanetary disk is substantially increased compared to an ideal gas polytrope when non-ideal effects such as hydrogen dissociation and ionization are taken into account. Moreover, our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

@@

One of the prevalent theories of giant planet formation is the core accretion model  (\citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}). In this model, a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere. In the standard core accretion models, the atmosphere is heated due to accretion of planetesimals, and as a result it radiates away energy. The envelope is therefore in a steady state at all times, and the atmosphere mass is a function of the core mass. It is thus found that there exists a minimum core mass past which hydrostatic equilibrium breaks down and rapid, unstable gas accretion commences: this is called the ``critical core mass''.

However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}). If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale. In Piso \& Youdin (in prep.) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion. We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk. We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope. We defined the time at which unstable atmosphere collapse commences as the \texit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities. We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for a higher mean molecular weight of the gas.

@@

 Piso \& Youdin (in prep.) assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$. In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account. A realistic hydrogen-helium mixture can be described using tabulated equation of state tables. In this study, we use the \cite{saumon95} EOS tables to describe the nebular gas. We generate atmosphere profiles and estimate atmospheric cooling timescales for a variety of disk conditions. From this, we determine the minimum core mass required for runaway gas accretion to commence within the typical life timescale of the protoplanetary disk.  We find that the realistic equation of state yields larger critical core masses compared to the ideal gas polytrope. %We show, however, that our results are still smaller than corresponding studies for large planetesimal accretion rates.

@@

This paper is organized as follows. In section \ref{sec2} we review the quasi-static and evolution models derived in Piso \& Youdin (in prep.). We discuss the variations in the adiabatic gradient, and hence in the atmosphere structure, caused by a non-ideal equation of state in section \ref{deladtable}, and discuss the implications of this variability on the atmosphere evolution time in section \ref{EOSeffects}. We determine the minimum core mass to form a giant planet during the disk lifetime when the nebular gas is described by a realistic equation of state in section \ref{critical}. In section \ref{acc} we compare our results to similar results obtained in studies that consider planetesimal accretion as the dominant source of energy. Finally, we summarize our findings in section \ref{conclusions}.

@@

In this section we review the model developed in Piso \& Youdin (in prep.) for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk. We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of a static atmosphere in section \ref{struct}.

@@

We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk. The two regions are separated by the Schwarzschild criterion for convective instability. We denote the surface between the two regions as the radiative-convective boundary (RCB), which is henceforth defined by a radius $r=R_{\cb}$. The evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion, and the luminosity is assumed to be constant throughout the outer radiative region. The atmosphere is considered to be spherically symmetric and self-gravitating. It consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively. We further assume that the envelope evolves through stages of quasi-static equilibrium.

@@

As a fiducial disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. In this model, the surface density and mid-plane temperature are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200 F_{\Sigma} a^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120 F_T a^{-3/7} \,\,K, \slabel{eq:diskb}
\end{subeqnarray}

\noindent with $a$ the semi-major axis in AU, and $F_{\Sigma}$ and $F_T$ normalization factors that adjust the disk mass and temperature relative to the minimum mass solar nebula (MMSN). In this study, we assume $F_{\Sigma}=F_T=1$. The resulting mid-plane pressure is given by

\begin{equation}
\label{eq:Pd}
P\di=110 F_{\Sigma} \sqrt{F_T m_*} a^{-45/14} \,\, \text{dyn cm}^{-2}
\end{equation}

\noindent for a mean molecular weight $\mu=2.35$. Here $m_* \equiv M_*/M_{\odot}$, where $M_*$ is the mass of the central star and $M_{\odot}$ is the mass of the Sun. We choose $m_*=1$.

%We further define several characteristic length scales that are important for this problem. \\
%
%\textbf{The Core Radius} represents the physical radius of the protoplanetary core:
%
%\begin{equation}
%\label{eq:rc}
%R_c \equiv \Big(\frac{3 M_c}{4 \pi \rho_c}\Big)^{1/3},
%\end{equation}
%
%\noindent with $M_c$ and $\rho_c$ the core mass and density, respectively. Since we assume that the core mass does not change with time, the core radius is also fixed for a given model. \\
%
%\textbf{The Bondi Radius} represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet. It is defined as
%
%\begin{equation}
%\label{eq:RB}
%R_B \equiv \frac{G M_p}{c_s^2}=\frac{G M_p}{\mathcal{R} T_d},
%\end{equation}
%
%\noindent where $G$ is the gravitational constant, $M_p$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. \\
%
%
%\textbf{The Hill Radius} represents the distance at which the gravitational attraction of the planet and of the host star are comparable. It is given by
%
%\begin{equation}
%\label{eq:RHill}
%R_H \equiv a \Big(\frac{M_p}{M_{\odot}}\Big)^{1/3}
%\end{equation}
%
%\noindent Outside the Hill radius, the gravity of the core is too weak to affect the nebular gas. Consequently, the planet will be able to accrete only gas that lies within its Hill sphere. \\
%
%\textbf{The Disk Scale Height} is defined by
%
%\begin{equation}
%\label{eq:H}
%H \equiv \frac{c_s^2}{\Omega},
%\end{equation}
%
%\noindent where $\Omega$ is the Keplerian angular velocity. The disk scale height is a measure of the thickness of the protoplanetary disk. Using equation (\ref{eq:diska}), we find that the scale height for our assumed disk model can be expressed as
%
%\begin{equation}
%\label{eq:H2}
%H=0.022 \sqrt{F_T} a^{9/7} \,\, \mathrm{AU}
%\end{equation}
%
%For small planet masses, $R_B<R_H<H$. As the atmosphere becomes more massive, tidal truncation becomes more important than thermal constraints, and $R_H<R_B<H$. Finally, for high mass atmospheres the disk scale height becomes smaller than the Hill radius of the planet, and spherical symmetry breaks, as the planet is now affected by the vertical profile of the disk.
%
%%Some previous studies of atmosphere accretion (e.g., \citealt{stevenson82}, \citealt{wuchterl93}, \citealt{rafikov06}) consider static envelopes, in which the luminosity is solely supplied by planetesimal accretion and fully radiated away by the atmosphere. In other studies, the time evolution is explicitly taken into account and full time dependent models are developed (e.g., \citealt{ikoma00}). We follow an intermediate approach and consider quasi static evolution. Our model for the atmosphere growth time is described in section \ref{...}.

%For the atmosphere structure, we develop a two-layer atmosphere model with the following assumptions:
%
%\begin{enumerate}
%\item The planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk, as mentioned above. The two regions are separated by the Schwarzschild criterion for convective instability (e.g., \citealt{thompson06}) 
%\item The luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction rather than planetesimal accretion. 
%\item The luminosity is assumed to be constant throughout the outer radiative region.
%\item The envelope evolves through stages of quasi-static equilibrium.
%\end{enumerate}

@@

An atmosphere in hydrostatic equilibrium is described by the following structure equations:

\begin{subeqnarray}
\label{eq:struct}
\frac{dP}{dr}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{dm}{dr}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{dT}{dr}&=&\nabla \frac{T}{P}\frac{dP}{dr}\slabel{eq:structc} \\
\frac{dL}{dr}&=&4 \pi r^2 \rho \epsilon_g\slabel{eq:structd}, 
\end{subeqnarray}

\noindent where $r$ is the radial coordinate, $P$, $T$ and $\rho$ are the gas pressure, temperature and density, respectively, $m$ is the mass enclosed by the radius $r$, $L$ is the luminosity from the surface of radius $r$, and $\epsilon_g \equiv -T \frac{ds}{dt}$ represents the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. The temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ has different expressions depending on the primary means of energy transport throughout the atmosphere. We assume that energy can be transported either through radiation or convection. When the luminosity is carried by radiative diffusion, the temperature gradient is given by

@@

\noindent where $\sigma$ is the Stefan-Boltzmann constant and $\kappa$ is the opacity. Alternatively, when the energy is transported outwards through convective motions, the temperature gradient becomes

@@

\noindent where $\delad$ is the adiabatic temperature gradient. The process that dominates energy transport throughout the atmosphere is determined by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when

\begin{equation}
\label{eq:schwarz}
\nabla < \delad
\end{equation}

\noindent and convectively unstable when the reverse is true. In order for convection to be effective, $\nabla \approx \delad$. Therefore, we find that the temperature gradient is given by $\nabla=\mathrm{min}(\delad, \delrad)$.

In order for equation set (\ref{eq:struct}) to be solvable, it has to be supplemented by an equation of state (EOS) that relates pressure, temperature and density $P=P(\rho, T)$, and an opacity law for $\kappa$. In this study we use the interpolated EOS tables of \citet{saumon95} for a helium fraction $Y=0.3$. More details on the EOS tables and the methodology of combining the separate tables for hydrogen and helium are presented in section \ref{deladtable} and \App{EOStables}.

We assume an opacity power law of the form

@@

\noindent with $\alpha$, $\beta$, $\kappa_0$ constants, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ a normalizing temperature and pressure, respectively. To estimate $\alpha$, $\beta$ and $\kappa_0$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $\kappa_0=2$. We note that these values are valid only for low disk temperatures: $T_d \lesssim 100 K$. As such, we only consider cool atmospheres that form in the outer part of the protoplanetary disk ($a \geq 5$ AU).

We now discuss our choice of core parameters and boundary conditions. We assume a solid core of fixed mass $M_c$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density. We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). Furthermore, we assume the atmosphere matches on to the protoplanetary at a distance equal to its Hill radius, the distance at which the gravitational attraction of the planet and of the host star are comparable. Its radius is given by

\begin{equation}
\label{eq:RHill}
R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}
\end{equation}

\noindent Outside the Hill radius, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet. We choose the effective outer boundary of the atmosphere to be the surface defined by the Bondi radius. The Bondi radius represents the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy binding the gas to the planet. It is defined as

\begin{equation}
\label{eq:RB}
R_{\rm B} \equiv \frac{G M_{\rm p}}{c_s^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di},
\end{equation}

\noindent where $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_s$ is the sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. Outside the Bondi sphere, the core gravity is too weak to significantly affect the nebular gas, justifying the choice of Bondi radius as the relevant scale for the atmosphere. We note that the Hill radius is still the correct scale for matching on to the disk, as perturbations of the nebular gas still occur past the Bondi radius due to the gravitational influence of the core. The above choice for the atmosphere boundary is only applicable when the Bondi radius is smaller than the Hill radius. If, on the other hand, $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$. For a given core mass, the atmosphere profile and evolution are therefore uniquely determined by the outer boundary conditions.

@@

Lastly, we review the cooling model developed in Piso \& Youdin (in prep.) used to determine the time evolution of the atmosphere between subsequent static models. A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:

@@

This cooling model applies at any radius $R$ where the mass enclosed is $M$, for example the Bondi radius, the Hill radius, or the radius of the radiative-convective boundary. $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay. $\Gamma$ is the heat generation rate, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a surface mass element.

As a consequence of the equations above, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. By connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}) we can obtain an evolutionary atmosphere series.

%\subsection{Quasi-Static Two-Layer Model}
%\label{twolayer}
%
%%The model and equations presented in this section so far can be applied to any protoplanetary atmosphere that satisfies the assumptions listed in section \ref{model}. 
%
%As a consequence of the equations derived in section \ref{cooling}, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth. In this section we therefore describe the procedure to obtain an evolutionary series from the cooling model introduced in \ref{cooling} between consecutive two-layer static atmospheres (cf. \ref{model}).
%
%We use the boundary conditions and opacity laws described in section \ref{struct}. The numerical integration is performed through the shooting method, which solves the boundary value problem by reducing it to an initial value problem: trial values are chosen for the parameters at one of the boundaries, then the equations are integrated and the resulting values at the other boundary are com- pared to the actual boundary conditions. The procedure is repeated until convergence is achieved. We start with a total atmosphere mass $M_i$, where the index $i$ labels each evolutionary stage. For this mass we guess a trial luminosity $L=L_{\mathrm{guess}}$. The assumption of constant luminosity throughout the radiative zone sets the right hand side of equation (\ref{eq:structb}) to zero, i.e. the time dependence is neglected. We integrate the structure equations (\ref{eq:struct}) inwards from $R_{\mathrm{out}}=R_H$, with the boundary conditions $T(R_{\mathrm{out}})=T_d$ and $P(R_{\mathrm{out}})=P_d$. The numerical integration gives a value for the core mass implied by the trial solution $M_c=M_{c, \mathrm{guess}}$. We adjust $L_{\mathrm{guess}}$ until $M_{c, \mathrm{guess}}$ converges to the actual core mass $M_c$. 
%
%The time evolution of the atmosphere is obtained using the global cooling model described in section \ref{cooling}. We neglect the luminosity due to planetesimal accretion and direct heat generation; as such, the first two terms in equation (\ref{eq:coolingglobal}) are set to zero. Equation (\ref{eq:coolingglobal}) is then evaluated at the RCB, since we assume that luminosity is primarily generated in the convective zone. In what follows, $X_{RCB}$ denotes quantity $X$ evaluated at the RCB. The time step $\Delta t_i$ between two consecutive models of masses $M_i$ and $M_{i+1}$ is calculated as follows:
%
%\begin{eqnarray}\nonumber
%\label{eq:dti}
%<L>_i \Delta t_i&=&-\Delta E_{RCB,i} + <e_{\mathrm{acc}, RCB}>_i \Delta M_{RCB, i} \\
%&&  - <P_{RCB}>_i \Delta V_{RCB, M, i} 
%\end{eqnarray}  
%
%\noindent where we denote 
%
%\begin{subeqnarray}
%\Delta X_i&=&X_{i+1}-X_i \\
%<X>_i&=&\frac{X_i+X_{i+1}}{2}
%\end{subeqnarray}
%\\
%\noindent The subscript $M$ in the volume expression above signifies evaluating the change in volume at constant mass. 
%
%By connecting sets of subsequent static atmospheres through the procedure described above we therefore obtain an evolutionary atmosphere series.

%

%\section{Quasi-Static Kelvin-Helmholtz Contraction}
%\label{KH}
%
%\textbf{Plots of radial / P-T profiles, also M-L, M-t profiles; compare the evolutionary profiles with the polytropic case. The text below is copy-pasted from Paper I - things will be similar, but needs a lot of rephrasing. }
%
%In this section we describe our choices for the gas, disk and core parameters, and present some typical atmosphere numerical models. As mentioned in section \ref{model}, we assume that the central star is a Sun-like star with $M_*=M_{\odot}$. We are primarily interested in the outer regions of the protoplanetary disk, so we generate atmosphere models between 10 and 100 AU. As a fiducial case, we assume that the atmosphere is composed of a hydrogen-helium mixture with the helium mass fraction $Y=0.3$., resulting in a mean molecular weight $\mu=2.35$. We do not take into account the dependence of the adiabatic index on the helium fraction, and therefore assume the adiabatic index for a diatomic gas $\delad=2/7$. We further explore how the atmosphere evolution varies with adiabatic index and molecular weight. We generate model atmospheres for monatomic adiabatic index $\delad=2/5$ and for a purely molecular hydrogen composition ($\mu=2$). The comparison between different gas compositions is deferred to section \ref{critcore}.
%

>>>>

\title{Quantitative Estimates of Minimum Core Masses for Giant Planet Formation}
\author{Ana-Maria A. Piso}
\affil{Harvard-Smithsonian Center for Astrophysics}
\author{Andrew N. Youdin}
\affil{JILA, University of Colorado at Boulder}
\author{Ruth A. Murray-Clay}
\affil{Harvard-Smithsonian Center for Astrophysics}

@@

The core accretion model assumes that giant planets form through gas accretion on to a solid core. The core and the atmosphere initially grow simultaneously through stages of quasistatic equilibrium. Once the core becomes massive enough, the atmosphere is no longer in hydrostatic balance and a rapid phase of runaway gas accretion commences. The minimum core mass for which unstable atmosphere collapse occurs is typically called the ``critical core mass''. In standard calculations of the critical core mass, planetesimal accretion dominates the atmosphere evolution, and the energy deposited by incoming planetesimal on to the core is radiated away by the atmosphere. In this study we consider a low planetesimal accretion regime in which the luminosity evolution of the atmosphere is dominated by Kelvin-Helmholtz contraction. We use the atmosphere structure and cooling model developed in Piso \& Youdin (2013) to derive the profile and evolution of atmospheres composed of a gas described by a realistic equation of state. We find that the minimum core mass (which we denote as \textit{critical core mass}) to form a giant planet before the dissipation of the protoplanetary disk is substantially increased compared to an ideal gas polytrope when non-ideal effects such as hydrogen dissociation and ionization are taken into account. Moreover, our results yield lower mass cores than corresponding studies for large planetesimal accretion rates. We therefore show that it is easier to form a planet by growing the core first, then accreting a massive gaseous envelope, rather than forming the core and atmosphere simultaneously.

@@

One of the prevalent theories of giant planet formation is the core accretion model  (e.g., \citealt{mizuno78}, \citealt{stevenson82}, \citealt{boden86}, \citealt{wuchterl93}, \citealt{dangelo11}). In this model, a solid core is first formed; the core grows, and once it becomes large enough it can accumulate a massive atmosphere. In standard core accretion models, in order to form a core large enough to attract an atmosphere, a high planetesimal accretion rate is needed, on average. The atmosphere is therefore heated due to accretion of planetesimals, and as a result it radiates away energy. The envelope is in a steady state at all times, and the atmosphere mass is a function of the core mass, which implies that every core mass maps uniquely to one atmosphere mass, for a given set of disk conditions and a planetesimal accretion rate. Moreover, larger cores hold fractionally larger envelopes. As a result, the atmosphere grows faster than the core, and eventually the atmosphere mass becomes comparable to the mass of the core. At this stage, hydrostatic equilibrium breaks down and a rapid phase of runaway gas accretion commences. This core mass $M\co$, for which the atmosphere mass $M_{\rm{atm}}$ satisfies $M_{\rm{atm}}\sim M\co$, is called the ``critical core mass'' in standard studies.

However, the planetesimal accretion rate is not necessarily constant at a given location in the protoplanetary disk throughout the disk lifetime (e.g., \citealt{ikoma00}). If the planetesimal accretion rate is very low, the atmosphere can no longer gain energy due to accretion of solids, and is instead dominated by gas accretion, contracting on a Kelvin-Helmholtz timescale. In Piso \& Youdin (2013; hereafter PY13) we studied the formation of giant planet atmospheres under the assumption that Kelvin-Helmoholtz gas contraction dominates the luminosity evolution of the atmosphere over planetesimal accretion. We built quasi-static two-layer atmosphere models with an inner convective region and an outer radiative region that matches smoothly onto the protoplanetary disk. We derived a cooling model to connect series of quasti-static atmospheres, and thus obtained an evolutionary history of the envelope. We defined the time $t$ at which unstable atmosphere collapse commences as the \textit{crossover time} $t_{\rm{co}}$, at which  $M_{\rm atm}(t_{\rm{co}})\sim M_{\rm c}$. From this we defined as \textit{critical core mass} the minimum core mass for a protoplanet to initiate runaway gas accretion during the lifetime of the protoplanetary disk. We studied this minimum mass for a variety of disk conditions, nebular gas compositions and opacities. We found that the critical core mass decreases for larger stellocentric distances, and is smaller for lower disk temperatures and opacities and for a higher mean molecular weight of the gas.

@@

 PY13 assume that the nebular gas can be described by a polytropic equation of state (EOS) corresponding to an ideal diatomic gas: $\delad=2/7$. In reality, however, non-ideal effects, such as gas dissociation and ionization, have to be taken into account. A realistic hydrogen-helium mixture can be described using tabulated equation of state tables. In this study, we use the \cite{saumon95} EOS tables to describe the nebular gas. We generate atmosphere profiles and estimate the atmospheric crossover time for a variety of disk conditions. From this, we determine the minimum core mass required for runaway gas accretion to commence within the typical life timescale of the protoplanetary disk.  We find that the realistic equation of state yields larger critical core masses compared to the ideal gas polytrope. %We show, however, that our results are still smaller than corresponding studies for large planetesimal accretion rates.

@@

This paper is organized as follows. In section \ref{sec2} we review the quasi-static and evolution models derived in PY13. We discuss the variations in the adiabatic gradient, and hence in the atmosphere structure, caused by a non-ideal equation of state in section \ref{deladtable}, and discuss the implications of this variability on the atmosphere evolution time in section \ref{EOSeffects}. We determine the minimum core mass to form a giant planet during the disk lifetime when the nebular gas is described by a realistic equation of state in section \ref{critical}. In section \ref{acc} we compare our results to similar results obtained in studies that consider planetesimal accretion as the dominant source of energy. Finally, we summarize our findings in section \ref{conclusions}.

@@

In this section we review the model developed in PY13 for the structure and evolution of a planetary atmosphere embedded in a protoplanetary disk. We describe the assumptions of the model and the properties of our assumed protoplanetary disk in section \ref{model}, and we summarize the equations describing the structure and time evolution of the atmosphere in section \ref{struct}.

@@

We assume that the planet consists of a solid core of fixed mass and a two-layer atmosphere composed of an inner convective zone and an outer radiative zone that matches smoothly on to the disk. The two regions are separated by the Schwarzschild criterion for convective instability. We denote the surface between the two regions as the radiative-convective boundary (RCB), defined by a radius $r=R_{\cb}$. We assume a low planetesimal accretion regime in which the atmosphere evolution is dominated by Kelvin-Helmholtz contraction while planetesimal accretion is neglected.  Moreover, we assume that the luminosity is constant throughout the outer radiative region. The atmosphere is spherically symmetric and self-gravitating. It consists of a hydrogen-helium mixture, with hydrogen and helium mass fractions of 0.7 and 0.3, respectively. We further assume that the envelope evolves through stages of quasi-static equilibrium.

@@

The temperature and pressure at the outer boundary of the atmosphere are given by the nebular temperature and pressure. As a disk model, we use the minimum mass, passively irradiated model of  \citet{chiang10}. The surface density, mid-plane temperature and mid-plane pressure are given by

\begin{subeqnarray}
\label{eq:diskparam}
\Sigma\di&=&2200\, (a/\text{AU})^{-3/2}\,\, \text{g cm}^{-2} \slabel{eq:diska}\\
T\di &=& 120\, (a/\text{AU})^{-3/7} \,\,K, \slabel{eq:diskb} \\
P\di&=&110\,  (a/\text{AU})^{-45/14} \,\, \text{dyn cm}^{-2} \slabel{eq:Pd}
\end{subeqnarray}

\noindent with $a$ the semi-major axis and for a mean molecular weight $\mu=2.35$.

@@

The structure of a static atmosphere is described by the standard equations of hydrostatic balance and thermal equilibrium:

\begin{subeqnarray}
\label{eq:struct}
\frac{d P}{d r}&=&-\frac{G m}{r^2}\rho \slabel{eq:structa} \\
\frac{d m}{d r}&=&4 \pi r^2 \rho\slabel{eq:structb} \\
\frac{d T}{d r}&=&\nabla \frac{T}{P}\frac{d P}{d r}\slabel{eq:structc} \\
\frac{d L}{d r}&=&4 \pi r^2 \rho (\epsilon + \epsilon_g)\slabel{eq:structd}, 
\end{subeqnarray}

\noindent with $r$ the radial coordinate, $P$, $T$ and $\rho$ the gas pressure, temperature and density, respectively, $m$ the mass enclosed by the radius $r$, $L$ the luminosity from the surface of radius $r$. The $\epsilon$ term represents the rate at which internal heat is generated per unit mass, while $\epsilon_g \equiv -T \frac{ds}{dt}$ is the heating per unit mass due to gravitational contraction, with $s$ the specific gas entropy. We do not take into account any internal energy sources and we therefore set $\epsilon=0$. The expression for the temperature gradient $\nabla \equiv \frac{d \ln T}{d \ln P}$ depends on whether energy is transported throughout the atmosphere by radiation or convection. In the case of radiative diffusion for an optically thick gas, the temperature gradient is

@@

\noindent with $\sigma$ the Stefan-Boltzmann constant and $\kappa$ the dust opacity. In our models the atmosphere is optically thick throughout the outer boundary. On the other hand, when the energy is transported outwards through convection, the temperature gradient is given by

@@

\noindent with $\delad$ the adiabatic temperature gradient. The convective and radiative layers of the envelope are separated by the Schwarzschild criterion (e.g., \citealt{thompson06}): the atmosphere is stable against convection when $\nabla < \delad$ and convectively unstable when $\nabla < \delad$. For effective convection, $\nabla \approx \delad$.  The temperature gradient is thus given by $\nabla=\mathrm{min}(\delad, \delrad)$.

The equation set (\ref{eq:struct}) is supplemented by an equation of state relating pressure, temperature and density, as well as an opacity law. In this study we use the interpolated EOS tables of \citet{saumon95} for a helium mass fraction $Y=0.3$. More details on the EOS tables and the methodology of combining the separate tables for hydrogen and helium are presented in section \ref{deladtable} and \App{EOStables}.

We assume a standard ISM opacity power law given by

@@

\noindent with $\alpha$, $\beta$, $\kappa_0$ constants, and $T_{\rm{ref}}$ and $P_{\rm{ref}}$ a normalizing temperature and pressure, respectively. To estimate $\alpha$, $\beta$ and $\kappa_0$ we use the \citet{bell94} opacity laws for ice grains: $\alpha =0 $, $\beta=2$ and $\kappa_0=2$. These values are valid only for low disk temperatures: $T_d \lesssim 100 K$.  We therefore consider protoplanets that form in the outer parts of the disk where temperatures are low, i.e. $a \geq 5$ AU. Moreover, dust settling or grain growth will result in different numerical values for the $\kappa_0$ and $\beta$ coefficients.

We further discuss our choice of core parameters and boundary conditions. We assume a solid core of fixed mass $M_{\rm c}$ with a radius $R_{\rm c}=(3 M_{\rm c}/4 \pi \rho_{\rm c})^{1/3}$, where $\rho_{\rm c}$ is the core density. We choose $\rho_{\rm c}=3.2$ g cm$^{-3}$ (e.g., \citealt{pap99}). The atmosphere is assumed to match on to the protoplanetary at the Hill radius, $R_{\rm H} \equiv a \Big(\frac{M_{\rm p}}{3 M_{\odot}}\Big)^{1/3}$, the distance at which the gravitational attraction of the planet and the tidal gravity due to the host star are equal. Outside the Hill sphere, the gravity of the planet is overcome by the tidal gravity from its host star, and hence only gas that lies within the Hill sphere can be gravitationally bound to the planet. The effective outer boundary of the atmosphere is the surface defined by the Bondi radius, $R_{\rm B} \equiv \frac{G M_{\rm p}}{c_{\rm s}^2}=\frac{G M_{\rm p}}{\mathcal{R} T\di}$. This is the distance from the planet at which the thermal energy of the nebular gas is of the order of the gravitational energy of the planet. Here $G$ is the gravitational constant, $M_{\rm p}$ is the total mass of the planet, $c_{\rm s}$ is the isothermal sound speed, and $\mathcal{R}$ is the reduced gas constant: $\mathcal{R}=k_b/(\mu m_p)$, with $k_b$ the Boltzmann constant and $m_p$ the proton mass. Outside the Bondi sphere, the gravity of the planet is too weak to significantly affect the nebular gas, which justifies the choice of Bondi radius as the relevant radius of the atmosphere. However, the nebular gas is still perturbed outside the Bondi sphere, and therefore the Hill radius is the correct scale for matching on to the disk. This choice for the atmosphere boundary applies only when the Bondi radius is smaller than the Hill radius; when $R_{\rm B}>R_{\rm H}$, the atmosphere only extends out to the Hill radius, since material cannot be gravitationally bound to the protoplanet outside the Hill sphere. At the Hill radius, the temperature and pressure are given by the nebular temperature and pressure: $T(R_{\rm H})=T_{\rm d}$ and $P(R_{\rm H})=P_{\rm d}$. For a given core mass, the atmosphere profile and evolution are therefore uniquely determined by the outer boundary conditions.

@@

Lastly, we review the cooling model developed in PY13 used to determine the time evolution of the atmosphere between subsequent static models. A protoplanetary atmosphere embedded in a gas disk satisfies the following cooling equation:

@@

Here, $L$ is the total luminosity, $L_{\rm c}$ is the luminosity from the solid core, and may include planetesimal accretion and radioactive decay, $\Gamma$ is the rate of internal heat generation, $\dot{E}$ is the rate at which total energy (internal and gravitational) is lost, and  $e_{\mathrm{acc}}$ is the specific total energy brought in by mass accreting at the rate $\dot{M}$: $e_{\mathrm{acc}}=u-G M/R$. The last term represents the work done on a surface mass element.

%As a consequence of the equations above, both the atmosphere structure and the gas accretion rate are uniquely determined by the current atmosphere mass. As this mass accretion rate is slow compared to the time it takes to relax to this solution, we can make a quasi-static model of the atmosphere growth.

As such, we can obtain an evolutionary series for the atmosphere by connecting sets of subsequent static atmospheres through the cooling equation (\ref{eq:coolingglobal}). Details of our numerical procedure are described in PY13.
